CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-06-20 script/de/chapters/primes.tex
*** empty log message ***

\textbf{Bewiesene Aussagen / S"atze zu Primzahlen}

\begin{itemize}
  \item Zu jeder Zahl $n$ aus ${\bf N}$ gibt es $n$ aufeinanderfolgende 
      nat"urliche Zahlen, die keine Primzahlen sind.
      Ein Beweis findet sich in \cite[S. 79]{Padberg1996}.
  \item Der ungarische Mathematiker Paul Erd"os (1913-1996) bewies:
      Zwischen jeder beliebigen Zahl ungleich $1$ und ihrem Doppelten gibt es
      mindestens eine Primzahl. Er bewies das Theorem nicht als erster, aber
      auf einfachere Weise als andere vor ihm.
      Gerade diese Vermutung legt nahe, dass wir auch heute noch nicht in
      aller Tiefe den Zusammenhang zwischen der Addition und der Multiplikation
      der nat"urlichen Zahlen verstanden haben.
  \item Es existiert eine reelle Zahl a, so dass die Funktion
      $f: {\bf N} \rightarrow {\mathbb Z}$ mit $n \mapsto a^{3^n}$
      f"ur alle $n$ nur Primzahlenwerte annimmt (siehe \cite[S. 82]
      {Padberg1996}).
      Leider macht die Bestimmung von $a$ Probleme (siehe unten).
\end{itemize}

\textbf{Unbewiesene Aussagen / Vermutungen zu Primzahlen}

\begin{itemize}
  \item Der deutsche Mathematiker \index{Goldbach Christian} Christian Goldbach (1690-1764) vermutete:
    Jede gerade nat"ur"-liche Zahl gr"o"ser $2$ l"asst sich als die Summe zweier Primzahlen darstellen.
    Mit Computern ist die Goldbachsche Vermutung f"ur alle geraden Zahlen bis $4*10^{14}$ verifiziert\footnote{%
Dass die Goldbachsche Vermutung wahr ist, d.h. f"ur alle geraden nat"urlichen Zahlen gr"o"ser als $2$ gilt, wird
    heute allgemein nicht mehr angezweifelt. Der Mathematiker J"org Richstein vom Institut f"ur Informatik der 
    Universit"at Gie"sen hat 1999 die geraden Zahlen bis 400 Billionen untersucht und kein Gegenbeispiel gefunden (siehe 
    \href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
         {\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).
    \index{Richstein 1999}
    Trotzdem ist das kein allgemeiner Beweis.\\ 
    Dass die Goldbach’sche Vermutung trotz aller Anstrengungen bis heute nicht bewiesen wurde, f"ordert 
    allerdings einen Verdacht:
    Seit den bahnbrechenden Arbeiten des "osterreichischen Mathematikers Kurt G"odel\index{G\""odel Kurt}
    ist bekannt, dass nicht jeder wahre Satz in der Mathematik auch beweisbar ist (Siehe 
    \href{http://www.mathematik.ch/mathematiker/goedel.html}
         {\tt http://www.mathematik.ch/mathematiker/goedel.html}).
    M"oglicherweise hat Goldbach also Recht,
    und trotzdem wird nie ein Beweis gefunden werden. Das wiederum l"asst sich aber vermutlich auch nicht beweisen.
    },   % footnote
    aber allgemein noch nicht bewiesen\footnote{%
Der englische Verlag {\em Faber} und die amerikanische
    Verlagsgesellschaft {\em Bloomsbury} publizierten 2000 das 1992 erstmal
    ver"offentlichte Buch \glqq Onkel Petros und die Goldbachsche Vermutung\grqq~ von Apostolos Doxiadis (deutsch bei
    L"ubbe 2000 und bei BLT als Taschenbuch 2001). Es ist 
    die Geschichte eines Mathematikprofessors, der daran scheitert, ein mehr als 250 Jahre altes R"atsel zu l"osen.\\
    Um die Verkaufszahlen zu f"ordern, schreiben die beiden  Verlage einen Preis von 1 Million USD aus, wenn 
    jemand die Vermutung beweist -- ver"offentlicht in einer angesehenen mathematischen Fachzeitschrift bis 2004 (Siehe 
    \href{http://www.mscs.dal.ca/~dilcher/Goldbach/index.html}
         {\tt http://www.mscs.dal.ca/\~{}dilcher/Goldbach/index.html}).\\
    Erstaunlicherweise d"urfen nur englische und amerikanische Mathematiker daran teilnehmen.\\
    Der Beweis, der der Goldbach-Vermutung bisher am n"achsten kommt, wurde 1966 von Chen Jing-Run
    bewiesen -- in einer schwer nachvollziehbaren Art und Weise: Jede gerade Zahl gr"o"ser 2 ist die Summe einer 
    Primzahl und des Produkts zweier Primzahlen. Z.B. $20=5+3*5.$\\ 
    Die wichtigsten Forschungsergebnisse zur Goldbachschen Vermutung sind zusammengefasst in dem von Wang Yuan 
    herausgegebenen Band: \glqq Goldbach Conjecture\grqq, 1984, 
    World Scientific Series in Pure Maths, Vol. 4.\\
    Gerade diese Vermutung legt nahe, dass wir auch heute noch nicht in aller Tiefe den Zusammenhang zwischen
    der Addition und der Multiplikation der nat"urlichen Zahlen verstanden haben. \\
    }.   % footnote
\item Der deutsche \index{Riemann Bernhard} Mathematiker Bernhard Riemann (1826-1866) stellte eine bisher 
    unbewiesene, aber auch nicht widerlegte (?) Formel f"ur die Verteilung von Primzahlen auf, die die Absch"atzung 
    weiter verbessern w"urde.
\end{itemize}

\textbf{Offene Fragestellungen} \\

Primzahlzwillinge sind Primzahlen, die den Abstand 2 voneinander haben, zum Beispiel 5 und 7 oder
101 und 103. Primzahldrillinge gibt es dagegen nur eines: 3, 5, 7.  Bei allen anderen Dreierpacks
aufeinanderfolgender ungerader Zahlen ist immer eine durch 3 teilbar und somit keine Primzahl.
\begin{itemize}
  \item Offen ist die Anzahl der Primzahlzwillinge: unendlich viele oder eine begrenzte Anzahl?
  Das bis heute bekannte gr"o"ste Primzahlzwillingspaar ist
  $1.693.965 \cdot 2^{66.443} \pm 1.$
  \item Gibt es eine Formel f"ur die Anzahl der Primzahlzwillinge pro Intervall?
  \item Der Beweis oben zu der Funktion  $f: \: N \rightarrow Z$ mit $n \mapsto a^{3^n}$  garantiert nur die Existenz einer
    solchen Zahl $a$.  Wie kann diese Zahl $a$ bestimmt werden, und wird sie einen Wert haben, so dass
    die Funktion auch von praktischem Interesse ist?
  \item Gibt es unendlich viele Mersenne-Primzahlen?\index{Primzahl!Mersenne}\index{Mersenne!Mersenne-Primzahl}
  \item Gibt es unendlich viele Fermatsche Primzahlen?
  \item Gibt es einen Polynomialzeit-Algorithmus\index{Polynom} zur 
       Zerlegung einer Zahl in ihre Primfaktoren (vgl. \cite[S. 167]{Klee1997})?
       Diese Frage kann man auf die beiden folgenden Fragestellungen
       aufsplitten:
  \begin{itemize}
        \item Gibt es einen Polynomialzeit-Algorithmus, der entscheidet, ob 
	      eine Zahl prim ist? \\
	      Diese Frage wurde durch den AKS-Algorithmus\index{AKS} 
	      beantwortet (vgl.  Kapitel \ref{PrimesinP}, ``Primes in P'':
	      Testen auf Primalit"at ist polynominal).
        \item Gibt es einen Polynomialzeit-Algorithmus, mit dem man feststellen
	      kann, aus wievielen Primfaktoren eine zusammengesetzte Zahl
	      besteht (ohne diese Primfaktoren zu berechnen)?
        \item Gibt es einen Polynomialzeit-Algorithmus, mit dem sich f"ur 
	      eine zusammengesetzte Zahl $n$ ein nicht-trivialer (d.h. 
	      von $1$ und von $n$ verschiedener) Teiler von $n$ berechnen 
	      l"asst?\footnote{Vergleiche auch Kapitel \ref{RSABernstein} 
	      und Kapitel \ref{NoteFactorisation}.}
        \end{itemize}
\end{itemize}

Am Ende von Kapitel \ref{NoteFactorisation} \hyperlink{RSA-160-chap3}{RSA-160}
k"onnen Sie die Gr"o"ßenordnungen ersehen, f"ur die heutige Algorithmen
Ergebnisse liefern (Primzahltests\index{Primzahltest}, 
Faktorisierung\index{Faktorisierung!Faktorisierungsrekorde}).

\paragraph{Weitere interessante Themen rund um Primzahlen}
\mbox{}
Nicht betrachtet wurden in diesem Kapitel folgende eher
zahlentheoretische Themen wie Teilbarkeitsregeln, Modulo-Rechnung,
modulare Inverse, modulare Potenzen und Wurzeln, chinesischer
Restesatz, Eulersche Phi-Funktion, perfekte Zahlen. Auf einige dieser
Themen geht das n"achste Kapitel ein.

>>>>

\paragraph{Bewiesene Aussagen / S"atze zu Primzahlen}
\begin{itemize}
  \item Zu jeder Zahl $n$ aus ${\bf N}$ gibt es $n$ aufeinanderfolgende 
      nat"urliche Zahlen, die keine Primzahlen sind.
      Ein Beweis findet sich in \cite[S. 79]{Padberg1996}.
  \item Der ungarische Mathematiker Paul Erd"os (1913-1996) bewies:
      Zwischen jeder beliebigen Zahl ungleich $1$ und ihrem Doppelten gibt es
      mindestens eine Primzahl. Er bewies das Theorem nicht als erster, aber
      auf einfachere Weise als andere vor ihm.
      Gerade diese Vermutung legt nahe, dass wir auch heute noch nicht in
      aller Tiefe den Zusammenhang zwischen der Addition und der Multiplikation
      der nat"urlichen Zahlen verstanden haben.
  \item Es existiert eine reelle Zahl a, so dass die Funktion
      $f: {\bf N} \rightarrow {\mathbb Z}$ mit $n \mapsto a^{3^n}$
      f"ur alle $n$ nur Primzahlenwerte annimmt (siehe \cite[S. 82]
      {Padberg1996}).
      Leider macht die Bestimmung von $a$ Probleme (siehe unten).
\end{itemize}

\paragraph{Unbewiesene Aussagen / Vermutungen zu Primzahlen}
\begin{itemize}
  \item Der deutsche Mathematiker \index{Goldbach Christian} Christian Goldbach (1690-1764) vermutete:
    Jede gerade nat"ur"-liche Zahl gr"o"ser $2$ l"asst sich als die Summe zweier Primzahlen darstellen.
    Mit Computern ist die Goldbachsche Vermutung f"ur alle geraden Zahlen bis $4*10^{14}$ verifiziert\footnote{%
Dass die Goldbachsche Vermutung wahr ist, d.h. f"ur alle geraden nat"urlichen Zahlen gr"o"ser als $2$ gilt, wird
    heute allgemein nicht mehr angezweifelt. Der Mathematiker J"org Richstein vom Institut f"ur Informatik der 
    Universit"at Gie"sen hat 1999 die geraden Zahlen bis 400 Billionen untersucht und kein Gegenbeispiel gefunden (siehe 
    \href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
         {\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).
    \index{Richstein 1999}
    Trotzdem ist das kein allgemeiner Beweis.\\ 
    Dass die Goldbach’sche Vermutung trotz aller Anstrengungen bis heute nicht bewiesen wurde, f"ordert 
    allerdings einen Verdacht:
    Seit den bahnbrechenden Arbeiten des "osterreichischen Mathematikers Kurt G"odel\index{G\""odel Kurt}
    ist bekannt, dass nicht jeder wahre Satz in der Mathematik auch beweisbar ist (Siehe 
    \href{http://www.mathematik.ch/mathematiker/goedel.html}
         {\tt http://www.mathematik.ch/mathematiker/goedel.html}).
    M"oglicherweise hat Goldbach also Recht,
    und trotzdem wird nie ein Beweis gefunden werden. Das wiederum l"asst sich aber vermutlich auch nicht beweisen.
    },   % footnote
    aber allgemein noch nicht bewiesen\footnote{%
Der englische Verlag {\em Faber} und die amerikanische
    Verlagsgesellschaft {\em Bloomsbury} publizierten 2000 das 1992 erstmal
    ver"offentlichte Buch \glqq Onkel Petros und die Goldbachsche Vermutung\grqq~ von Apostolos Doxiadis (deutsch bei
    L"ubbe 2000 und bei BLT als Taschenbuch 2001). Es ist 
    die Geschichte eines Mathematikprofessors, der daran scheitert, ein mehr als 250 Jahre altes R"atsel zu l"osen.\\
    Um die Verkaufszahlen zu f"ordern, schreiben die beiden  Verlage einen Preis von 1 Million USD aus, wenn 
    jemand die Vermutung beweist -- ver"offentlicht in einer angesehenen mathematischen Fachzeitschrift bis 2004 (Siehe 
    \href{http://www.mscs.dal.ca/~dilcher/Goldbach/index.html}
         {\tt http://www.mscs.dal.ca/\~{}dilcher/Goldbach/index.html}).\\
    Erstaunlicherweise d"urfen nur englische und amerikanische Mathematiker daran teilnehmen.\\
    Der Beweis, der der Goldbach-Vermutung bisher am n"achsten kommt, wurde 1966 von Chen Jing-Run
    bewiesen -- in einer schwer nachvollziehbaren Art und Weise: Jede gerade Zahl gr"o"ser 2 ist die Summe einer 
    Primzahl und des Produkts zweier Primzahlen. Z.B. $20=5+3*5.$\\ 
    Die wichtigsten Forschungsergebnisse zur Goldbachschen Vermutung sind zusammengefasst in dem von Wang Yuan 
    herausgegebenen Band: \glqq Goldbach Conjecture\grqq, 1984, 
    World Scientific Series in Pure Maths, Vol. 4.\\
    Gerade diese Vermutung legt nahe, dass wir auch heute noch nicht in aller Tiefe den Zusammenhang zwischen
    der Addition und der Multiplikation der nat"urlichen Zahlen verstanden haben. \\
    }.   % footnote
\item Der deutsche \index{Riemann Bernhard} Mathematiker Bernhard Riemann (1826-1866) stellte eine bisher 
    unbewiesene, aber auch nicht widerlegte (?) Formel f"ur die Verteilung von Primzahlen auf, die die Absch"atzung 
    weiter verbessern w"urde.
\end{itemize}

\paragraph{Offene Fragestellungen}\mbox{}\\
Primzahlzwillinge sind Primzahlen, die den Abstand 2 voneinander haben, zum Beispiel 5 und 7 oder
101 und 103. Primzahldrillinge gibt es dagegen nur eines: 3, 5, 7.  Bei allen anderen Dreierpacks
aufeinanderfolgender ungerader Zahlen ist immer eine durch 3 teilbar und somit keine Primzahl.
\begin{itemize}
  \item Offen ist die Anzahl der Primzahlzwillinge: unendlich viele oder eine begrenzte Anzahl?
  Das bis heute bekannte gr"o"ste Primzahlzwillingspaar ist
  $1.693.965 \cdot 2^{66.443} \pm 1.$
  \item Gibt es eine Formel f"ur die Anzahl der Primzahlzwillinge pro Intervall?
  \item Der Beweis oben zu der Funktion  $f: \: N \rightarrow Z$ mit $n \mapsto a^{3^n}$  garantiert nur die Existenz einer
    solchen Zahl $a$.  Wie kann diese Zahl $a$ bestimmt werden, und wird sie einen Wert haben, so dass
    die Funktion auch von praktischem Interesse ist?
  \item Gibt es unendlich viele Mersenne-Primzahlen?\index{Primzahl!Mersenne}\index{Mersenne!Mersenne-Primzahl}
  \item Gibt es unendlich viele Fermatsche Primzahlen?
  \item Gibt es einen Polynomialzeit-Algorithmus\index{Polynom} zur 
       Zerlegung einer Zahl in ihre Primfaktoren (vgl. \cite[S. 167]{Klee1997})?
       Diese Frage kann man auf die folgenden drei Fragestellungen
       aufsplitten:
  \begin{itemize}
        \item Gibt es einen Polynomialzeit-Algorithmus, der entscheidet, ob 
	      eine Zahl prim ist? \\
	      Diese Frage wurde durch den AKS-Algorithmus\index{AKS} 
	      beantwortet (vgl.  Kapitel \ref{PrimesinP}, ``Primes in P'':
	      Testen auf Primalit"at ist polynominal).
        \item Gibt es einen Polynomialzeit-Algorithmus, mit dem man feststellen
	      kann, aus wievielen Primfaktoren eine zusammengesetzte Zahl
	      besteht (ohne diese Primfaktoren zu berechnen)?
        \item Gibt es einen Polynomialzeit-Algorithmus, mit dem sich f"ur 
	      eine zusammengesetzte Zahl $n$ ein nicht-trivialer (d.h. 
	      von $1$ und von $n$ verschiedener) Teiler von $n$ berechnen 
	      l"asst?\footnote{Vergleiche auch Kapitel \ref{RSABernstein} 
	      und Kapitel \ref{NoteFactorisation}.}
        \end{itemize}
\end{itemize}

Am Ende von Kapitel \ref{NoteFactorisation}, 
Abschnitt \hyperlink{RSA-160-chap3}{RSA-160}
k"onnen Sie die Gr"o"senordnungen ersehen, f"ur die heutige Algorithmen
Ergebnisse liefern (Primzahltests\index{Primzahltest}, 
Faktorisierung\index{Faktorisierung!Faktorisierungsrekorde}).

\paragraph{Weitere interessante Themen rund um Primzahlen}
\mbox{}\\
Nicht betrachtet wurden in diesem Kapitel folgende eher
zahlentheoretische Themen wie Teilbarkeitsregeln, Modulo-Rechnung,
modulare Inverse, modulare Potenzen und Wurzeln, chinesischer
Restesatz, Eulersche Phi-Funktion, perfekte Zahlen. Auf einige dieser
Themen geht das n"achste Kapitel ein.

### 2003-06-20 script/en/chapters/introduction.tex
be updates
*** empty log message ***

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes, ``M-40'') 
	\\ (chap. \ref{spezialzahlentypen}, \ref{zahlentyp_mersenne}), 
  \item progress in number theory (``Primes in P'') (chap. \ref{PrimesinP}), 
  \item the factorisation of big numbers (TWIRL, RSA-160)\index{TWIRL device} 
        (chap. \ref{TWIRLDevice}, \ref{RSA-160}),
  \item progress in cryptanalysis of the AES standard 
        (chap. \ref{NeueAES-Analyse}) and
  \item progress in brute-force attacks against symmetric algorithms 
        (chap. \ref{Brute-force-gegen-Symmetr}).
\end{itemize}

@@

I am deeply grateful to all the people helping with their impressive
commitment who have made this global project so successful. Especially I would
like to acknowledge the English language proof-reading of this script version
done by Richard Christenson and L. Montgomery.

>>>>

This edition of the script mainly updates the summaries of the following 
topical research areas:
\vspace{-7pt}
\begin{itemize}
  \item the search for the largest prime numbers (generalized Mersenne 
        and Fermat primes) \\ 
%        and Fermat primes, ``M-40'') 
	(chap. \ref{spezialzahlentypen}, \ref{zahlentyp_mersenne}), 
  \item progress in number theory (``Primes in P'') (chap. \ref{PrimesinP}), 
  \item the factorisation of big numbers (TWIRL, RSA-160)\index{TWIRL device} 
        (chap. \ref{TWIRLDevice}, \ref{RSA-160}),
  \item progress in cryptanalysis of the AES standard 
        (chap. \ref{NeueAES-Analyse}) and
  \item progress in brute-force attacks against symmetric algorithms 
        (chap. \ref{Brute-force-gegen-Symmetr}).
\end{itemize}

@@

I am deeply grateful to all the people helping with their impressive
commitment who have made this global project so successful. Especially I would
like to acknowledge the English language proof-reading of this script version
done by Richard Christensen and Lowell Montgomery.

### 2003-06-20 script/en/chapters/primes.tex
*** empty log message ***

% --------------------------------------------------------------------------
\subsection{Notes}

\textbf{Proven statements / theorems about primes}

\begin{itemize}
  \item For each number $n$ in ${\bf N}$ there are $n$ consecutive natural
      numbers that are not primes.
      A proof of this can be found in \cite[p. 79]{Padberg1996}.
  \item The Hungarian mathematician Paul Erd\"os (1913-1996) proved:
      Between each random number not equal to $1$ and its double, there is
      at least one prime. He was not the first to prove this theorem, but
      proved it in a much simpler manner than those before him.
  \item There is a real number a such that the function $f: {\bf N}
      \rightarrow {\mathbb Z}$ where $n \mapsto a^{3^n}$ 
      only delivers primes for all $n$ (see \cite[p. 82]{Padberg1996}).
      Unfortunately, problems arise when we try to determine $a$ (see below).
\end{itemize}

\textbf{Unproven statements / conjectures about primes}

\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
      (1690-1764) conjectured: Every even natural number greater than 2 can
      be represented as the sum of two prime numbers.
      Computers have verified\footnote{%
It is generally accepted today, that the Goldbach Conjecture is true, 
i.~e.\ valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richstein from the computer sciences 
institute at the University of Giessen, studied even numbers up to 
400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).
\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact is that despite all efforts, Goldbach's conjecture has to date
not been proven. This leads one to believe that since the pioneer work of
the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is 
well-known, not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}
{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbach's conjecture was correct, but in any case the proof
will never be found. Conversely, that will presumably also remain unproven.}
      the Goldbach conjecture for all even numbers up to
      $4*10^{14}$ but no general proof has yet been found\footnote{%
The English publisher {\em Faber} and the American publisher {\em Bloomsbury}
issued in 2000 the 1992 published book ``Uncle Petros and Goldbach's 
Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than
250 year old puzzle. To boost the sales figures the English and American
publishers have offered a prize of 1 million USD, if someone can prove
the conjecture -- which should be published by 2004 in a well-known
mathematical journal.\\
Surprisingly only British and American citizens are allowed to participate.\\
The theorem which has come closest so far to Goldbach's conjecture
was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
to understand:
Each even integer greater than $2$ is the sum of one prime and of 
the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in 
the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes it clear, that even today we do not
have a complete understanding of the deeper connections between addition and
multiplication of natural numbers.}.

@@

\textbf{Open questions} \\

Twin primes are prime numbers whose difference is 2. Examples include 5 and 7 or
101 and 103. Triplet primes, however, only occur once: 3, 5, 7.  For all other
sets of three consecutive uneven numbers, one of them is always divisible by 3
and thus not a prime.
\begin{itemize}
  \item The number of twin primes is an open question: infinite or limited
number?
  The largest twin primes known today are $1,693,965 \cdot 2^{66,443} \pm 1.$
  \item Does a formula exist for calculating the number of twin primes per
interval?
  \item The above proof of the function  $f: \: N \rightarrow Z$ with $n \mapsto
a^{3^n}$  only guarantees the existence of such a number $a$.  How can we
determine this number $a$ and will it have a value, making the function also of
some practical interest?
\item Is there an infinite number of Mersenne prime numbers?\index{Prime number!Mersenne}\index{Mersenne!prime number}
\item Is there an infinite number of Fermat prime numbers?

  \item Does a polynomial\index{Polynomial} time algorithm exist for 
        calculating the prime factors of a number (see \cite[p. 167]{Klee1997})?
	This question can be divided into the two following questions:
   \begin{itemize} 
      \item Does a polynomial time algorithm exist that decides whether a
            number is prime? \\
	    This question has been answered by the AKS algorithm\index{AKS} 
	    (see chapter \ref{PrimesinP}, ``Primes in P'': 
	    Primality testing is polynomial).

      \item Does a polynomial time algorithm exist  that calculates for a
            composite number fom how many prime factors it is made up 
	    (without calculating these factors)?

@@

At the end of chapter \ref{NoteFactorisation} \hyperlink{RSA-160-chap3}{RSA-160}
you can see the dimensions for which numbers current algorithms deliver
results (test for primality\index{Primality testing}, 
factorization\index{Factorisation!factoring records}).

\paragraph{Further interesting topics regarding prime numbers}
\mbox{}

This chapter doesn't consider other number theory topics such as
divisibility rules, modulus calculation, modular inverses, modular powers and
roots, Chinese remainder theorem, Euler PHI function, perfect numbers.
Some of these are considered in the next chapter.

>>>>

% --------------------------------------------------------------------------
\pagebreak
\subsection{Notes}
\paragraph{Proven statements / theorems about primes}
\begin{itemize}
  \item For each number $n$ in ${\bf N}$ there are $n$ consecutive natural
      numbers that are not primes.
      A proof of this can be found in \cite[p. 79]{Padberg1996}.
  \item The Hungarian mathematician Paul Erd\"os (1913-1996) proved:
      Between each random number not equal to $1$ and its double, there is
      at least one prime. He was not the first to prove this theorem, but
      proved it in a much simpler manner than those before him.
  \item There is a real number a such that the function $f: {\bf N}
      \rightarrow {\mathbb Z}$ where $n \mapsto a^{3^n}$ 
      only delivers primes for all $n$ (see \cite[p. 82]{Padberg1996}).
      Unfortunately, problems arise when we try to determine $a$ (see below).
\end{itemize}

\paragraph{Unproven statements / conjectures about primes}\mbox{}
\begin{itemize}
  \item The German mathematician \index{Goldbach Christian} Christian Goldbach
      (1690-1764) conjectured: Every even natural number greater than 2 can
      be represented as the sum of two prime numbers.
      Computers have verified\footnote{%
It is generally accepted today, that the Goldbach Conjecture is true, 
i.~e.\ valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richstein from the computer sciences 
institute at the University of Giessen, studied even numbers up to 
400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).
\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact is that despite all efforts, Goldbach's conjecture has to date
not been proven. This leads one to believe that since the pioneer work of
the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is 
well-known, not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}
{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbach's conjecture was correct, but in any case the proof
will never be found. Conversely, that will presumably also remain unproven.}
      the Goldbach conjecture for all even numbers up to
      $4*10^{14}$ but no general proof has yet been found\footnote{%
The English publisher {\em Faber} and the American publisher {\em Bloomsbury}
issued in 2000 the 1992 published book ``Uncle Petros and Goldbach's 
Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than
250 year old puzzle. To boost the sales figures the English and American
publishers have offered a prize of 1 million USD, if someone can prove
the conjecture -- which should be published by 2004 in a well-known
mathematical journal.\\
Surprisingly only British and American citizens are allowed to participate.\\
The theorem which has come closest so far to Goldbach's conjecture
was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
to understand:
Each even integer greater than $2$ is the sum of one prime and of 
the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in 
the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes it clear, that even today we do not
have a complete understanding of the deeper connections between addition and
multiplication of natural numbers.}.

@@

\paragraph{Open questions}\mbox{}\\
Twin primes are prime numbers whose difference is 2. Examples include 5 and 7 or
101 and 103. Triplet primes, however, only occur once: 3, 5, 7.  For all other
sets of three consecutive uneven numbers, one of them is always divisible by 3
and thus not a prime.
\begin{itemize}
  \item The number of twin primes is an open question: infinite or limited
number?
  The largest twin primes known today are $1,693,965 \cdot 2^{66,443} \pm 1.$
  \item Does a formula exist for calculating the number of twin primes per
interval?
  \item The above proof of the function  $f: \: N \rightarrow Z$ with $n \mapsto
a^{3^n}$  only guarantees the existence of such a number $a$.  How can we
determine this number $a$ and will it have a value, making the function also of
some practical interest?
\item Is there an infinite number of Mersenne prime numbers?\index{Prime number!Mersenne}\index{Mersenne!prime number}
\item Is there an infinite number of Fermat prime numbers?

  \item Does a polynomial\index{Polynomial} time algorithm exist for 
        calculating the prime factors of a number (see \cite[p. 167]{Klee1997})?
	This question can be divided into the three following questions:
   \begin{itemize} 
      \item Does a polynomial time algorithm exist that decides whether a
            number is prime? \\
	    This question has been answered by the AKS algorithm\index{AKS} 
	    (see chapter \ref{PrimesinP}, ``Primes in P'': 
	    Primality testing is polynomial).

      \item Does a polynomial time algorithm exist  that calculates for a
            composite number from how many prime factors it is made up 
	    (without calculating these factors)?

@@

At the end of chapter \ref{NoteFactorisation},
section \hyperlink{RSA-160-chap3}{RSA-160}
you can see the dimensions for which numbers the current algorithms deliver
results (test for primality\index{Primality testing}, 
factorization\index{Factorisation!factoring records}).

\paragraph{Further interesting topics regarding prime numbers}
\mbox{}\\
This chapter doesn't consider other number theory topics such as
divisibility rules, modulus calculation, modular inverses, modular powers and
roots, Chinese remainder theorem, Euler PHI function, perfect numbers.
Some of these are considered in the next chapter.
